Imputation of Incomplete Multilevel Data with \proglang{R}

### 2023-11-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd

title:
  formatted: "Imputation of Incomplete Multilevel Data with \\proglang{R}"
  plain:     "Imputation of Incomplete Multilevel Data with R"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}. Our scope is only simple multilevel models, to show how imputation can yield less biased estimates from incomplete clustered data. More complex models can be accomodated, but are outside the scope of this paper. 
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

| **Concept**         | **Details**                                                                   |
|---------------------|-------------------------------------------------------------------------------|
| Sample unit         | Units of the population from which measurements are taken in a sample.        |
| Hierarchical levels | Data are grouped into clusters at different levels. A three-level             |
| Fixed effect        | Here we assume that the values of an independent variable are fixed, i.e.,    |
|                     | the values observed in the study are representative of all values in the      |
|                     | in the dependent variable y e.g., blood pressure between treatments A and B.  |
| Random effect       | The values of an independent variable are assumed to be randomly drawn from   |
|                     | admission we might select only certain hospitals that are representative of   |
|                     | the difference of y between individual hospitals, but rather the variation of |
| ICC                 | The variability due to clustering is often measured by means of the           |
|                     | intraclass coefficient (ICC). The ICC can be seen as the percentage           |
|                     | of variance that can be attributed to the cluster-level, where a high         |
|                     | ICC would indicate that a lot of variability is due to the cluster            |
|                     | structure.                                                                    |
| Random effect       | Multilevel models typically accommodate for variability by including          |
|                     | a separate group mean for each cluster. In addition to random                 |
|                     | intercepts, multilevel models can also include random coefficients            |
|                     | and heterogeneous residual error variances across clusters [see e.g.          |
|                     | @gelm06, @hox17 and @jong21]. [TODO: add stratified intercept as concept..]                     |

@@

We'll now use `school` as a predictor to impute all other variables. This is still not recommended practice, since it only works under certain circumstances and results may be biased [@drec15; @ende16]. But at least, it includes some multilevel aspect. This method is also called 'fixed cluster imputation', and uses N-1 indicator variables representing allocation of N clusters as a fixed factor in the model [@reit06; @ende16]. Colloquially, this is 'multilevel imputation for dummies'.

@@

Impute the incomplete data
```{r imp_impact, cache=TRUE}
imp <- mice(dat, method = meth, predictorMatrix = pred, printFlag = FALSE)
```

@@

Analyze the imputed data:
```{r impact-fit, cache=TRUE}
fit <- imp %>% 
  with(glmer(
    mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name),
    family = "binomial"
    )) 
# tidy(pool(fit))
# as.mitml.result(fit)
# testEstimates(as.mitml.result(fit))
```
The estimated effects after imputation are presented in Table XYZ.

@@

@@

>>>>

title:
  formatted: "Imputation of Incomplete Multilevel Data with \\proglang{R}"
  plain:     "Imputation of Incomplete Multilevel Data with R"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}. Our scope is only simple multilevel models, to show how imputation can yield less biased estimates from incomplete clustered data. More complex models can be accomodated, but are outside the scope of this paper. Incomplete multilevel data requires careful consideration of the missing data problem and analysis strategy. In this tutorial, we focus on a popular strategy for accommodating missingness in multilevel data: replacing the missing data with one or more plausible values, i.e., imputation.Imputation separates the missing data problem from the main analysis and the completed data can be analyzed as if it has been fully observed.This tutorial illustrates the imputation of incomplete multilevel data with the statistical programming language R. We aim to show how imputation can yield less biased estimates from incomplete clustered data. We provide practical guidelines and code snippets for different missing data situations, including non-ignorable missingness mechanisms. For brevity, we focus on multilevel imputation using chained equations with the R mice package and its adjacent packages.
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

```{r clus0,  echo = FALSE, warning = FALSE,cache = TRUE}
concepts <- data.frame(Concept=c("Sample units","Cluster","Hierarchical data","Level-1","Level-2","Hierarchical model","Fixed effect","Random effect","Mixed effect","ICC","Stratified intercept"),
                       Details=c("Units of the population from which measurements are taken in a sample, e.g., students.", "Variable that specify the cluster or agruppation, e.g., Classroom","Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.","Variable that varies within a cluster, eg. Test score","Variable that does not vary within a cluster but between, e.g. teacher experience.","Model accounting for dependant observations relying on certain parameters ( within cluster) which in turn depend on other parameters (between cluster)" ,"Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as  a teaching strategy (tutoring after class)","Effects that are a source of random variation in the data, and whose levels are not fully sampled. e.g. test score tendency during academic year between students due to no controlled factors such as  genetic,family history","Includes fixed and random effects, e.g. the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated. Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals. In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g. @gelm06, @hox17 and @jong21).","The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage
of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.",
                                 ""))
library(kableExtra)
concepts%>%
  kable( booktabs = T, escape = F,
         caption = "Concepts in multilevel methods",
         col.names = c("Concept","Details"), align = "l")%>%
  column_spec(c(1), width = "3cm")%>%
  column_spec(c(2), width = "12cm")
```

@@

We will now use `school` as a predictor to impute all other variables. This is still not recommended practice, since it only works under certain circumstances and results may be biased [@drec15; @ende16]. But at least, it includes some multilevel aspect. This method is also called 'fixed cluster imputation', and uses N-1 indicator variables representing allocation of N clusters as a fixed factor in the model [@reit06; @ende16]. Colloquially, this is 'multilevel imputation for dummies'.

@@

Impute the incomplete data
```{r imp_impact}
imp <- mice(dat, method = meth, predictorMatrix = pred, printFlag = FALSE)
```

@@

Analyze the imputed data:
```{r impact-fit}
fit <- imp %>% 
  with(glmer(
    mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name),
    family = "binomial"
    )) 
# tidy(pool(fit))
# as.mitml.result(fit)
# testEstimates(as.mitml.result(fit))
```
The estimated effects after imputation are presented in Table XYZ.

@@

In this example, we demonstrate a multilevel imputation of random intercept and random slope model with a continuous response. We utilize the obesity dataset included in the `micemd`\@ package, a simulated dataset that emulates an electronic survey in which individuals are asked to provide information about their weight and consumption habits in different countries. We simulate data for 5 clusters so that the true values are known. We use the following variables from the dataset:

-   **Cluster:** Region of the patients' healthcare provider (Cluster variable),
-   **Gender:** Subjects' Gender (0=male, 1=female),
-   **Age:** Subjects' age,
-   **Height:** Subjects' height in metres,
-   **Weight:** Subjects' weight in kilograms,
-   **BMI:** Subjects' body mass index,
-   **FamOb:** Family obesity history (yes or no),
-   **Time:** Response time in minutes (exclusion-restriction variable).

In this dataset, Age and FamOb are MAR, while the weight variable is affected by selection bias, attributed to an indirect MNAR mechanism. This MNAR mechanism typically arises when an unobserved or omitted variable influences both the value of the incomplete variable (in this case, Weight) and its likelihood of being missing (denoted as R).

In the primary analysis model, BMI serves as the dependent variable, with Age, Gender, and FamOb as predictors. Because of the clustered nature of the data, which is quantified with the Intraclass Correlation Coefficient (ICC) below, we include random intercepts, as well as a random slope for the Age variable.
The model is represented as:
\begin{equation}
\label{eqn:main}
BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FamOb_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}
\end{equation}

```{r, include=FALSE, eval=FALSE}
library(here)
source(here("heckman","ipdfunction.R"),local =knitr::knit_global())
load(here("heckman","Obesity.rda"))
```

We start by loading the data:

```{r}
#data("data_heckman", package = "micemd")
#dat <- data_heckman
```

Now, let's begin by examining the missing patterns in the data by cluster:

```{r, dpi = 300,out.width="70%", fig.cap="Missing pattern", eval=FALSE}
library(ggpubr)
myplots <- lapply(1:5, function(i) {
  ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+
  ggplot2::ggtitle(paste0("Cluster", i))
 })
ggarrange(myplots[[1]], myplots[[3]], nrow=1,common.legend = TRUE, legend="bottom")
```

We observe that the missing pattern is non-monotonic and quite similar across the clusters. However, regarding the weight variable, we notice that is systematically missing in cluster 3. In order to evaluate
if we require a imputation method that accounts for clustering  we assess the Intraclass Correlation

```{r obesity-icc, eval=FALSE}
Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
#Performance::ICC(Nulmodel)
```

Since the ICC is above 0.1 and as the main analysis will be use a mixed model, we decide
to use two-level (2l) imputation methods. In this imputation process, we include all predictor
variables from equation \ref{eqn:main} in the main model. However, since BMI is a composite of
weight and height, we use deterministic imputation for these, which is described below.

We use the **find.defaultMethodfunction** provided in the **micemd** package, which suggests an appropriate method for MAR variables based on the type of variable, number of observations in the cluster, and number of clusters.

It suggests using '2l.2stage.bin' for the FAV variable and '2l.2stage.norm' for the age variable. However, after inspecting the age density plot, we consider modifying its method to
'2l.2stage.pmm'. For the BMI variable, we employ deterministic imputation.

```{r obesitymeth, eval=FALSE}
meth_mar <- find.defaultMethod(Obesity, ind.clust=1, I.small = 7,
                               ni.small = 100, prop.small = 0.4)
meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
meth_mar["Age"]<-"2l.2stage.pmm" 
```

For these imputation models, it is necessary to specify the prediction matrix, with the cluster
variable labelled as -2 and the predictor variable measured within clusters labelled as 2, encompassing all variables. We need to suprime the variable Time as this variable is not specified in the main model. We also  modify the relationship between BMI, weight and height in the prediction matrix to avoid circular predictions. Then we proceed to run the imputation model.
```{r obesity-predmar,out.width="70%", eval=FALSE}
pred_mar <- quickpred(Obesity)   # predictor matrix
pred_mar[,"Cluster"] <- -2 # clustering variable
pred_mar[,"Time"] <- 0
pred_mar[pred_mar==1] <- 2
pred_mar[c("Height", "Weight"), "BMI"] <- 0
ggmice::plot_pred(pred_mar)
imp_mar <- mice(data = Obesity, meth = meth_mar, pred = pred_mar,
                m=10, seed = 123, printFlag = FALSE)
```

```{r obesity-predmar1,eval=FALSE}
summary(complete(imp_mar,"long")$Weight)
```
We are also contemplating the utilisation of the ???pmm??? option, as the values imputed using a fully parametric method may be implausibly low for some patients.

```{r obesity-predmar_pmm, eval=FALSE}
meth_mar["Weight"]<-"2l.2stage.pmm" 
imp_mar_pmm <- mice(data = Obesity, meth = meth_mar, pred = pred_mar,
                    m=10, seed = 123, printFlag = FALSE)
```

```{r obesity-predmar_pmm1,eval=FALSE}
summary(complete(imp_mar_pmm,"long")$Weight)
#ggmice::ggplot_trace(imp_mar_pmm, "Weight")
```

After confirming convergence, we proceed to save the results for future use.  We consider the possibility that patients may not have been selected randomly, which would then have led to a distribution for weight that does not reflect the weight in the population. It???s likely that an omitted variable, like self-esteem, could influence this selection. For instance, individuals with lower self-esteem might have higher weight values, impacting their willingness to provide honest information due to embarrassment.

To address this situation, two approaches have been proposed for dealing with Missing Not at Random (MNAR) data: pattern-mixed models and selection models. Within pattern-mixed models, methods like the delta method and more advanced ones like NARFS have been suggested. The selection model approach includes methods such as the Heckman model, which can be particularly useful in this case. Several methods, including those by \cite{Galimar_2017,Hammon_2021}, and the recently a Heckman method designed for two-level data, allow for variations in intercepts and exposure effects (random intercept and slope).

To apply the **2l.2stage.heckman** method, the weight variable should be specified as '2l.2stage.heckman' found in the micemd package. Additionally, the prediction matrix needs modification because this method involves specifying two equations: one for the outcome, describing the incomplete variable in terms of partially observed predictors (in this case, all variables from the main model), and the other for the selection model, explaining the probability of being observed based (R) on certain variables.
For the outcome equation we consider the same imputation model that we used for
the MAR case (main model).
 $$Weight_{ij}= \beta^O_{o} + \beta^O_{1}Age_{ij} + \beta^O_{2}FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}$$ 
Regarding the selection equation, we include the same predictors as those in the main model, as well as a time variable. Here the time variable serves as a restriction exclusion variable specifically explaining the probability of being observed but not affecting the incomplete value (Weight). In this context, we assume that the time a user spends completing the survey serves as a proxy for the barriers they may encounter in survey completion, such as familiarity with the survey content or internet speed. These factors may lead the user to skip specific questions or even the entire survey. Also, we assume the time does not have any influence on the subject???s weight.
 $$R_{ij}= \beta^S_{o} + \beta^S_{1}Age_{ij} + \beta^S_{2}FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}$$

These two equations are jointly estimated under the assumption that the error terms are interconnected with a bivariate normal distribution. For a more comprehensive understanding of the model and the exclusion restriction, see \cite{Munoz_2022}.

To use information from both equations, we must adjust the prediction matrix. The cluster variable remains specified as before (-2). In this imputation method, all the variables present in both the selection and outcome equations are included with a random effect.

However, it is essential to distinguish which of these variables appear in each equation. In this framework, when a variable is shared between both equations, it is denoted as (2). Predictors exclusive to the outcome equation are indicated as (-4), while those exclusive to the selection equation are labelled as (-3). Consequently, the only alteration needed in the predictor matrix pertains to the variable 'Time'.

```{r obesity-predmnar, eval=FALSE}
pred_mnar <- pred_mar
pred_mnar["Weight","Time"]<- -3
#ggmice::plot_pred(pred_mnar)
```

We also need to modify the method of the weight variable.

```{r obesity-predmnar1, eval=FALSE}
meth_mnar <- meth_mar
meth_mnar["Weight"]<- "3l.2stage.heckman"
```
Then we proceed to run the imputation model as before, after executing these imputation procedures, it is essential to assess convergence and the coherence of the imputed values.
```{r obesity-predmnar2,eval=FALSE}
imp_mnar<- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
                m=10, seed = 123, printFlag = FALSE)
summary(complete(imp_mnar,"long")$Weight)
```

Upon examining the weight variable, we noticed that the imputed range falls outside the realm of plausible values (as weight should be positive).

```{r obesity-predmnar3,eval=FALSE}
summary(complete(imp_mnar,"long")$Weight)
```
Consequently, as before we  use the 'pmm', option but this time for the Heckman imputation, this approach ensures that the imputed values remain within the range of observable values. We then run the imputation model but this time using the option of pmm, to assure that weight values are in the range of the observable data, this can be implemented by setting the pmm parameter to true.

```{r obesity-predmnarp, eval=FALSE}
imp_mnar_pmm <- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
                     m=10, seed = 123, pmm = T,  printFlag = FALSE)
```

We check the convergency of the results
```{r obesity-predmnarp1, eval=FALSE}
summary(complete(imp_mnar_pmm,"long")$Weight)
#ggmice::ggplot_trace(imp_mar_pmm, "Weight")
```

After this modification we proceed to compare the effects on the model. We
run the analysis model on each of the completed datasets as well asthe dataset where the
incomplete values are removed (Complete Case analysis, CC).

```{r models, eval=FALSE}
library(ggplot2)
cc_rs<- with(setDT(Obesity)[complete.cases(Obesity),],
             lme( BMI ~ Age + FamOb + Gender, random=~1+Age|Cluster))
mar_rs <- with(imp_mar,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
mar_pmm_rs <- with(imp_mar_pmm,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
mnar_rs<- with(imp_mnar,lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
mnar_pmm_rs<- with(imp_mnar_pmm, lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
plot_models(list_models,
            mod_name = c("Complete case", "MAR","MAR_pmm", "MNAR", "MNAR_pmm"))
```

We note that there is minimal disparity in the age effect, FamObs, or  Gender across the various imputation models under consideration. An analysis of the intercept reveals that, under the MNAR assumption, a higher average BMI is anticipated compared to the MAR assumption.
Nonetheless, with respect to precision of estimates, we notice that in general MNAR imputation leads to wider confidence intervals, in this case it does not have any influence on the final result but there could be cases where variation in the assumed missing mechanism could lead also to differences on significant test and therefore lead to contradictory conclusions.

# Additional

Visualize missing data pattern:

@@

ORDER:

-   summary

-   congeniality, then in hierarchical models

-   look whether we can fit cong. back in the main body

-   alt. methods

-   conclusion: mice is really easy!
